---
title: "Assignment 4"
author: "Ignacio Lafuente"
date: "10/2/2020"
output:
  html_document:
    toc: true
    toc_float: true
    highlight: tango
    code_folding: hide
---

## Creative Assignment 4 - Lines and Networks

### Loading libraries and data

My analysis will be about 'relevant points of my life', i.e. places were I have lived, studied or worked within the city of Buenos Aires, and their accessibility by foot and by bicycle.

```{r load libraries, message=FALSE, results='hide'}
library(osmdata)
library(opentripplanner)
library(tidyverse)
library(sf)
library(ggthemes)
library(ggspatial)
library(RColorBrewer)
```

```{r importing files, message=FALSE, results='hide'}
points <- read.csv2("ba_points.csv")
```

### Setting coordinates projection system

```{r projecting coordinates}
campoinchauspe <- "+proj=tmerc +lat_0=-34.629717 +lon_0=-58.4627 +k=1 +x_0=100000 +y_0=100000 +a=6378388 +b=6378386.996621622 +towgs84=0,0,0,0,0,0,0 +units=m +no_defs"

points2 <- st_as_sf(points, coords = c("X", "Y"), crs = campoinchauspe)
```

### Loading street networks for Buenos Aires

```{r}
ba_street_query <- opq(bbox = 'Buenos Aires Argentina') %>%
  add_osm_feature(key = 'highway')

ba_street_query %>%
  osmdata_xml(file = 'OTP/graphs/default/ba_streets.osm')

ba_street_features <- ba_street_query %>%
  osmdata_sf()

ba_streets <- ba_street_features$osm_lines %>%
  st_transform(crs = campoinchauspe)

```

### Network preview

```{r}
ggplot(ba_streets) +
  geom_sf() +
  theme_map()
```
### Setting up Open Trip Planner

```{r, message=FALSE, results='hide'}
path_data <- file.path(getwd(), "OTP")
path_otp <- paste(path_data, "otp.jar",sep = "/")
otp_build_graph(otp = path_otp, dir = path_data, memory = 1024) 
```

```{r}
otp_setup(otp = path_otp, dir = path_data, memory =1024)
otpcon <- otp_connect()
```

### Creating and plotting isochrones

I will be creating isochrones for a seven-minute period, by foot and by bicycle. Five minutes seemed short for a relevant walking distance in Buenos Aires, but ten minutes led to big sized isochrones which were overlapping one another. That is why I chose this odd number.

```{r generating isochrones, results = 'hide', message = FALSE}
iso_7min_walk <- 
  otp_isochrone(otpcon = otpcon, fromPlace = points2,                 mode = "WALK", cutoffSec = 420) %>%
  st_transform(crs = campoinchauspe) %>%
  mutate(mode = "walk")
iso_7min_bike <- 
  otp_isochrone(otpcon = otpcon, fromPlace = points2,
                mode = "BICYCLE", cutoffSec = 420) %>%
  st_transform(crs = campoinchauspe) %>%
  mutate(mode = "bicycle")
iso_all_modes <- rbind(iso_7min_bike, iso_7min_walk)

otp_stop()
```

```{r plotting isochrones}
right_side <- st_bbox(iso_all_modes)$xmax
left_side  <- st_bbox(iso_all_modes)$xmin
top_side <- st_bbox(iso_all_modes)$ymax
bottom_side <- st_bbox(iso_all_modes)$ymin

ggplot(iso_all_modes) +
  annotation_map_tile(zoomin = 0, progress = "none") +
  geom_sf(aes(fill = mode), alpha = 0.4) +
  geom_sf(data = points2) + ## not showing
  coord_sf(xlim = c(left_side, right_side), 
           ylim = c(bottom_side, top_side), expand = FALSE) +
  scale_fill_brewer(name = "Area that is reachable within 7 minutes",
                    palette = "Set2",
                    labels = c("By bicycle", "By foot")) +
  theme_map() +
  labs(caption = "Basemap Copyright OpenStreetMap contributors")
```

### Spatial join (keeping points' records in the isochrones)

```{r}
iso_all_modes2 <- st_join(iso_all_modes, points2) ##empty output
```


### Analyzing relations between isochrones

```{r}
iso_areas <- iso_all_modes %>%
  mutate(area = st_area(iso_all_modes)) %>%
  st_set_geometry(NULL) %>%
  pivot_wider(names_from = mode, values_from = area) %>%
  mutate(ratio_btow = as.numeric(bicycle) / as.numeric(walk))
```

### Defining scalars for Euclidean distance buffers

I am assuming average walking speed is 5 km/h, whereas the average biking speed is 13 km/h (this was my frequent commute, taking some red lights into account, for example). The analysis might be biased if OTP used substantially different standards.

These average speeds times the seven-minute period would be the distance, and thus the radii of a buffer. The area of that buffer would be, under some assumptions, TT * r^2.

```{r scalars euclidean}
bike = pi * (7/60 * 13)^2
walk = pi * (7/60 * 5)^2
```

#### figures: 45° line (implicit ratio btow), ratio btow, ratio isochrone to euclidean for both modes