---
title: "Assignment 1"
author: "Ignacio Lafuente"
date: "9/13/2020"
output: html_document
---
# Creative Assignment 1

## Relationships and distributions

### Selected county:
Montgomery County, MD

### Selected variables:
* Rural (dummy)
* Majority ethnicity
* Percentage of renter households
* Average household size
* Total population
* Median age

(observations are census tracts)

```{r load libraries, results='hide', message=FALSE}
library(ggplot2)
library(tidyverse)
library(tidycensus)
library(viridis)
library(ggcorrplot)
```

```{r select variables}
vars <- c(tot_hh = "H001001",
          tot_rural_hh = "H002005",
          tot_rent_hh = "H004004",
          avg_hh_size = "H012001",
          tot_pop = "P003001",
          nh_white_pop = "P005003",
          nh_black_pop = "P005004",
          nh_asian_pop = "P005006",
          hispanic_pop = "P005010",
          med_age = "P013001")

montg_tracts <- get_decennial(geography = "tract", variables = vars,
                           state = "MD", county = "Montgomery county",
                           output = "wide") %>%
  mutate(rural = (tot_rural_hh == tot_hh) ,
         pct_rental = tot_rent_hh / tot_hh,
         majority_race_eth = case_when(
           nh_white_pop / tot_pop > 0.5 ~ "nh_white",
           nh_black_pop / tot_pop > 0.5 ~ "nh_black",
           nh_asian_pop / tot_pop > 0.5 ~ "nh_asian",
           hispanic_pop / tot_pop > 0.5 ~ "hispanic",
           (nh_white_pop + nh_black_pop + nh_asian_pop + hispanic_pop) /
             tot_pop < 0.5 ~ "other",
           TRUE ~ "None")) %>%
  filter(tot_pop > 0, tot_hh > 0) %>%
  select(GEOID, rural, majority_race_eth, 
         pct_rental, avg_hh_size, tot_pop, med_age)
```

## Plot 1

### Trends for median age and percentage of renters correlation, according to majority ethnicity

```{r trends for median age and rent by majority ethnicity}
ggplot(montg_tracts, 
       aes(x = pct_rental,
           y = med_age,
           color = majority_race_eth)) +
  geom_point() +
  stat_smooth(method = "lm", se = FALSE) +
  scale_x_continuous(name = "Percent of households in rented homes",
                     breaks = seq(0, 1, by = 0.20),
                     labels = paste(seq(0, 100, by = 20), "%", sep = "")) +
  scale_y_continuous(name = "Median age within tract",
                     breaks = seq(10, 80, by = 10)) +
  scale_color_discrete(name = "Majority race/ethnicity",
                       labels = c("Hispanic",
                                  "Non-Hispanic Black",
                                  "Non-Hispanic White",
                                  "No majority")) +
  theme_bw() 
```

## Plot 2

### Average household size by majority ethnicity
```{r average household size by majority ethnicity}
ggplot(montg_tracts, 
       aes(x = majority_race_eth,
           y = avg_hh_size,
           alpha = tot_pop)) +
  geom_point(position = "jitter",
             size = 0.5) +
  scale_x_discrete(name = "Majority race/ethnicity",                        labels = c("Hispanic",
                                  "Non-Hispanic Black",
                                  "Non-Hispanic White",
                                  "No majority")) +
  scale_y_continuous(name = "Average household size within tract",
                     breaks = seq(1, 8, by = 1)) +
  theme(panel.background = element_rect(fill = "white"),
        panel.grid = element_line(colour = "gray",
        linetype = "dotted"),
        legend.key = element_rect(fill = "white"))
```

## Plot 3

### Number of tracts of each majority ethnicity
```{r bar chart}
ggplot(montg_tracts, aes(x = majority_race_eth, fill = rural)) +
  geom_bar(color = "blue", fill = "white") +
  scale_y_continuous(name = "Number of tracts") +
  scale_x_discrete(name = "Majority race/ethnicity",
                       labels = c("Hispanic",
                                  "Non-Hispanic\nBlack",
                                  "Non-Hispanic\nWhite",
                                  "No majority")) +
   theme_bw()
```

## Plot 4

### Median age according to urban/rural type of tract
```{r median age according to urban/rural}
ggplot(montg_tracts, 
       aes(x = rural,
           y = med_age)) +
  geom_boxplot() +
  scale_x_discrete(name = "Tract type",
                       labels = c("Urban",
                                  "Rural")) +
  scale_y_continuous(name = "Median age within tract",
                     breaks = seq(10, 80, by = 10)) +
  theme(panel.background = element_rect(fill = "white"),
        panel.grid = element_line(colour = "gray",
        linetype = "dashed")) +
  theme(legend.position = "none")
```

## Plot 5

### Pie chart of tract types
```{r pie chart about urban rural}
pie_chart <- ggplot(montg_tracts, aes(x = 1, fill = rural)) +
  geom_bar(position = "fill") +
  coord_polar(theta = "y") +
  scale_y_continuous(name = "Number of tracts") +
  scale_fill_discrete(name = "Tract type",
                       labels = c("Urban",
                                  "Rural")) +
  theme_void()

pie_chart
```

## Plot 6

### Density of total population for census tracts
```{r density of something}
ggplot(montg_tracts, aes(x=tot_pop)) +
  geom_density(fill="gray", color="black", size=1.5, alpha=0.8) +
  scale_x_continuous(name = "Total population of tract",
                     breaks = seq(0, 10000, by = 1000)) +
  theme_minimal()
```

## Plot 7

### Median age by majority ethnicity
```{r median age by majority ethnicity}
ggplot(montg_tracts, aes(x=majority_race_eth, y=med_age)) +
  geom_point() + 
  geom_segment(aes(x=majority_race_eth, xend=majority_race_eth, y=0, yend=med_age)) + 
  scale_y_continuous(name = "Median age within tract",
                     breaks = seq(0, 80, by = 10)) +
  scale_x_discrete(name = "Majority race/ethnicity",                       labels = c("Hispanic",
                                  "Non-H Black",
                                  "Non-H White",
                                  "No Majority")) +

theme_bw()
```

## Plot 8

### Quadratic fit for correlation between percentage of renters and total population of tract
```{r total population and total rent percentage}
ggplot(montg_tracts, aes(x=tot_pop, y=pct_rental))+
  geom_point(alpha = 0.5) +
  stat_smooth(method = "lm", formula = y ~ poly(x,2), se = FALSE, colour = "orange") +
  scale_y_continuous(name="Percentage of households who are renters",
                     breaks = seq(0, 1, by = 0.1),
                     labels = paste(seq(0, 100, by = 10), "%", sep = "")) +
  scale_x_continuous(name="Total population of tract") +
  theme_classic()
```

## Plot 9

### Logarithmic fit for correlation between median age and average household size
```{r age and household size}
ggplot(montg_tracts,
       aes(x=med_age, y=avg_hh_size)) +
  geom_point(colour = "light blue") +
  stat_smooth(method="lm",formula = y ~ log(x)) +
  scale_y_continuous(name="Average household size within tract") +
  scale_x_continuous(name="Median age within tract")+
  theme_dark()
```

## Plot 10

### Correlogram for some continuous variables:
#### Average household size, total population and median age
```{r corrplot}
vars2 <- c(avg_hh_size = "H012001",
          tot_pop = "P003001",
          med_age = "P013001")
montg_tracts2 <- get_decennial(geography = "tract", variables = vars2, state = "MD", county = "Montgomery county", output = "wide") %>%
  filter(tot_pop > 0) %>%
  select(avg_hh_size, tot_pop, med_age)
corr <- round(cor(montg_tracts2), 1)
ggcorrplot(corr, hc.order = TRUE, 
           type = "upper", 
           lab = TRUE, 
           lab_size = 3, 
           method="square", 
           colors = c("tomato2", "white", "springgreen3"),
           ggtheme=theme_bw)

```