---
title: "Assignment 2"
author: "Ignacio Lafuente"
date: "9/20/2020"
output: html_document
---

### Projections and Distortions

```{r load libraries, message = FALSE, results = 'hide'}
library(ggplot2)
library(sf)
library(rnaturalearth)
library(tidyverse)
library(ggspatial)
library(cartogram)
library(ggthemes)
library(dplyr)
library(wesanderson)
library(RColorBrewer)
```

```{r world map}
world <- ne_countries(scale = 50, returnclass = "sf") 
world_map <- ggplot(world) +
  geom_sf() +
  theme_classic()

world_map
```

## South America default map

```{r South America}
sa_countries <- world %>%
  filter(continent == "South America")

sa_map <- ggplot(sa_countries) +
  geom_sf() + 
  theme_bw() +
  annotation_scale()

sa_map
```

## Projected Coordinate Systems

##### 1 EPSG:3114 MAGNA-SIRGAS / Colombia Far West zone
##### 2 SR-ORG:8347 Lambert Azimutal Equal Area Bahia Brazil
##### 3 EPSG:2081 Chos Malal 1914 / Argentina zone 2

My intention was to pick opposite extreme points (far North-West, far East, far Sorth)

```{r projected coordinate systems}
col_fwest <- "+proj=tmerc +lat_0=4.596200416666666 +lon_0=-80.07750791666666 +k=1 +x_0=1000000 +y_0=1000000 +ellps=GRS80 +towgs84=0,0,0,0,0,0,0 +units=m +no_defs"
bahia_br <- "+proj=laea +lat_0=-13 +lon_0=-42 +x_0=600000 +y_0=600000 +ellps=GRS80 +units=m +no_defs"
patagonia_ar <- "+proj=tmerc +lat_0=-90 +lon_0=-69 +k=1 +x_0=2500000 +y_0=0 +ellps=intl +units=m +no_defs"
```

## South America - Colombia Far West Zone projection

```{r South America proj 1}
ggplot(sa_countries) +
  geom_sf() + 
  coord_sf(crs = col_fwest) +
  theme_bw() +
  annotation_scale()
```

## South America - Lambert Azimutal Equal Area Bahia Brazil projection

```{r South America proj 2}
ggplot(sa_countries) +
  geom_sf() + 
  coord_sf(crs = bahia_br) +
  theme_bw() +
  annotation_scale()
```

## South America - Chos Malal 1914 / Argentina Zone 2 projection

```{r South America proj 3}
ggplot(sa_countries) +
  geom_sf() + 
  coord_sf(crs = patagonia_ar) +
  theme_bw() +
  annotation_scale()
```

```{r transform}
countries_transformed <- sa_countries %>%
  st_transform(crs = patagonia_ar)
```

#### Importing Covid-19 data

```{r importing covid deaths, message = FALSE, results = 'hide'}
covid_world <- read_csv("https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_daily_reports/09-20-2020.csv")

covid_world_2 <- select(covid_world, Country_Region, Deaths)

covid_world_3 <- aggregate(covid_world_2[-1], by = list(covid_world$Country_Region), FUN = sum)

covid_sa <- countries_transformed %>%
  left_join(covid_world_3, by = c("name" = "Group.1")) %>%
  select(name, pop_est, Deaths)

covid_sa$deaths_per_m <- covid_sa$Deaths / covid_sa$pop_est * 1000000

```

## Covid accumulated deaths by country, per million inhabitants, as of 09-20-2020

```{r deaths per million inhabitants}
ggplot(covid_sa, aes(fill = deaths_per_m)) +
  geom_sf(color = NA) +
  scale_fill_viridis_c(
    name = "Number of\nreported\nCOVID-19\naccumulated\ndeaths per million\ninhabitants as of\nSeptember 20, 2020",
    breaks = seq(0, 1000, by = 200),
    labels = formatC(seq(0, 1000, by = 200), 
                     format = "f", digits = 0)) +
  theme_map() +
  theme(legend.background = element_blank(),
        legend.position="left")
```

A cloropleth of (in my opinion) the most relevant metric about the Covid-19 pandemic.

#### Importing area data and defining population density and GDP per capita

```{r population density, message = FALSE, results = 'hide'}
world_area <- read_csv2("areas_sa.csv")

sa_countries_2 <- countries_transformed %>%
  left_join(world_area, by = c("name" = "country")) %>%
  select(name, pop_est, tot_area_km2, gdp_md_est)

sa_countries_2$density <- sa_countries_2$pop_est / sa_countries_2$tot_area_km2

sa_countries_2$y <- sa_countries_2$gdp_md_est / sa_countries_2$pop_est * 1000000
```

## Continuous Cartogram - GDP per capita

(Outdated)

```{r mapping density, message = FALSE, results = 'hide'}
gdppc_cartogram_cont <- sa_countries_2 %>%
  cartogram_cont("y")

ggplot(gdppc_cartogram_cont) +
  geom_sf(fill = "#330066", color = "gray") +
  theme_map()
```

Even if the aesthetics could be much better, I personally do not find these cartograms useful, besides the curiosity of understanding their underlying logic for the first time. I think a lot of communication efficiency is lost when the shapes people are used to see are not available.

## Proportional symbol map on Population Density

```{r proportional symbol map, message = FALSE, results = 'hide'}
sa_centeroids <- sa_countries_2 %>%
  st_centroid()

ggplot(sa_countries_2) +
  geom_sf(fill = "#CCCCCC", color = "black") +
  geom_sf(data = sa_centeroids, 
          aes(size = density),
          alpha = 0.5, color = "#990000") +
  scale_size_continuous(name = "Population density\nas inhabitants per km2",
    breaks = seq(0, 60, by = 10),
    labels = formatC(seq(0, 60, by = 10), 
                     format = "f", digits = 0),
    range = c(0, 20)) +
  theme_void() 
```

I like this format. The only issue is that plotting bigger circles or a larger range would increase the legend size since it is shown 1:1, and the map would seem irrelevant.
